`conmat`: generate synthetic contact matrices for a given age-stratified population

### 2025-01-20 paper/paper.qmd
suggested paper edits

Contact matrices describe the number of contacts between individuals. They are used to create models of infectious disease spread. `conmat` is an R package which generates synthetic contact matrices for arbitrary input demography, ready for use in infectious diseases modelling.

@@

Infectious diseases like influenza and COVID19 spread via social contact. If we can understand patterns of contact---which individuals are more likely be in contact with each other---then we will be able to create models of how disease spreads. Epidemiologists and public policy makers can use these models to make decisions to keep a population safe and healthy.

@@

A prominent contact survey is the "POLYMOD" study by @mossong2008, which surveyed 8 European countries: Belgium, Germany, Finland, Great Britain, Italy, Luxembourg, The Netherlands, and Poland [@mossong2008].

These social contact surveys can be projected on to a given demographic structure to produce estimated daily contact rates between age groups. These are known as "contact matrices" or "synthetic contact matrices". A widely used approach by @prem2017 [@prem2021] produced contact matrices for 177 countries at "urban" and "rural" levels for each country.

However, there were major limitations with the methods in @prem2021. First, not all countries were included in their analyses. Second, the contact matrices only covered broad scale areas. This presents challenges for decision makers who are often working at a sub-national geographical scale. Third, the code provided by Prem et al. was not designed for replicability and easy modification with user-defined inputs.

The `conmat` package was developed to fill the specific need of creating contact matrices for arbitrary age categories and populations (as shown in the below example). We developed the method primarily to output synthetic contact matrices. We also provided methods to create next generation matrices.

@@

We can then generate a contact matrix for Tasmania, from the POLYMOD study with `extrapolate_polymod()`.

@@

```{r}
#| label: fig-autoplot-contacts
#| fig-cap: "Contact patterns between individuals for different age groups across four settings: home, work, school, and other. The x axis shows the age group from, and the y axis shows the age group to, coloured by the average number of contacts between individuals in two age groups. We see different contact patterns in different settings, for example, diagonal with wings for the home setting."
#| fig-width: 8
#| fig-height: 8
autoplot(tasmania_contact)
```

@@

1) fit a model to predict individual contact rate, using an existing contact survey;
2) predict a synthetic contact matrix using age population data.

@@

These covariates capture typical features of inter-person contact, where individuals primarily interact with people of similar age (the diagonals of the matrix), and with grandparents and/or children (the so-called 'wings' of the matrix). The key features of the relationship between the age groups, represented by the six covariates, are displayed in @fig-show-partial-plots for the home setting. The $|i-j|$ term gives the strong diagonal, modelling people generally living with those of similar age, and the $\max(i,j)$ and $\min(i,j)$ terms give the intergenerational effect of parents and grandparents with children.

@@

We provide a trained model of contact rate that is fit to the POLYMOD survey for ease of use. The software also has an interface to use other contact surveys, such as @comix. This is important as POLYMOD represents contact patterns in 8 countries in Europe, and contact patterns are known to differ across nations and cultures.

@@

This work was used as a key input into several models for COVID-19 transmission and control in Australia and contributed to decisions around vaccination policy [@DohertyModelling].

@@

* demonstrate fitting a model to other contact surveys from sources such as the `socialmixr` R package [@socialmixr];
* predict to any age brackets - such as monthly ages, for example, 1, 3, 6, month year old infants;
* fitting to multiple contact surveys simultaneously, e.g., POLYMOD and CoMix;
* add uncertainty to estimates;
* add methods for including household size distributions;
* include known household age distributions as offsets in the 'home' setting model, in place of the whole population distribution.

>>>>

Contact matrices describe the number of contacts between individuals. They are used to create models of infectious disease spread. `conmat` is an R package which generates synthetic contact matrices for arbitrary input demography, ready for use in infectious disease modelling.

@@

Infectious diseases like influenza and COVID-19 spread via social contact. If we can understand patterns of contact---which individuals are more likely be in contact with each other---then we will be able to create models of how disease spreads. Epidemiologists and public policy makers can use these models to make decisions to keep a population safe and healthy.

@@

A prominent contact survey is the POLYMOD study by @mossong2008, which surveyed 8 European countries: Belgium, Germany, Finland, Great Britain, Italy, Luxembourg, The Netherlands, and Poland [@mossong2008].

These social contact surveys can be projected on to a given demographic structure to produce estimated daily contact rates between age groups. These are known as 'synthetic' contact matrices. A widely used approach by @prem2017 [@prem2021] produced synthetic contact matrices for 177 countries at 'urban' and 'rural' levels for each country.

However, there were major limitations with the methods in @prem2021. First, not all countries were included in their analyses. Second, the contact matrices only covered broad population groups within entire countries. This presents challenges for decision makers who are often working at a sub-national geographical scale, with differing demographic structure in different sub-populations. Third, the code provided by Prem et al. was not designed for replicability and easy modification with user-defined inputs.

The `conmat` package was developed to fill the specific need of creating contact matrices for arbitrary age categories and populations (as shown in the below example) to inform infectious diease models. We developed the method primarily to output synthetic contact matrices. We also provided methods to create next generation matrices for modelling.

@@

We can then generate a synthetic contact matrix for Tasmania, by extrapolating the contact patterns between age groups learned from the POLYMOD study, using `extrapolate_polymod()`.

@@

```{r}
#| label: fig-autoplot-contacts
#| fig-cap: "Contact patterns between individuals for different age groups across four settings: home, work, school, and other. The x axis shows the age groups for focal individuals ('from'), and the y axis shows the age groups for people those individuals have contact with ('to'), coloured by the average number of contacts the individual has in that age group. We see different contact patterns in different settings, for example, diagonal with 'wings' for the home setting."
#| fig-width: 8
#| fig-height: 8
autoplot(tasmania_contact)
```

@@

1) fit a model to predict individual contact rates, using an existing contact survey;
2) predict a synthetic contact matrix using age population data.

@@

These covariates capture typical features of inter-person contact, where individuals primarily interact with people of similar age (the diagonals of the matrix), and with grandparents and/or children (the so-called 'wings' of the matrix). The key features of the relationship between the age groups, represented by spline transformations of the six covariates, are displayed in @fig-show-partial-plots for the home setting. The spline-transformed $|i-j|$ and $|i-j|^{2}$ terms give the strong diagonal lines modelling people generally living with those of similar age and the intergenerational effect of parents and (faintly) grandparents with children. The spline-transformed $\text{max}(i, j)$ and $\text{min}(i, j)$ terms give the higher rates of at-home contact among younger people of similar ages and with their parents.

@@

We provide a trained model of contact rate that is fit to the POLYMOD survey for ease of use. The software also has an interface to train models to other contact surveys, such as @comix. This is important as POLYMOD represents contact patterns in 8 countries in Europe, and contact patterns are known to differ across nations and cultures.

@@

This code underlying this software was used as a key input into several models for COVID-19 transmission and control in Australia and contributed to decisions around vaccination policy [@DohertyModelling].

@@

* integrating model fitting methods with contact survey prepared using the `socialmixr` R package [@socialmixr];
* enabling prediction to arbitrary age brackets, e.g. monthly age bins for infants;
* fitting to multiple contact surveys simultaneously, e.g., POLYMOD and CoMix;
* providing estimates of uncertainty in contact matrices;
* adding methods to include household size distributions in predictions of contacts in the 'home' setting;
